Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-04 _tex/paper2.tex
Built site for gh-pages

Luckily, SCMs and DAGs support identification analysis through two key
advantages. First, regardless of complexity, they can represent various
causal structures using only five fundamental building blocks. This
feature allows researchers to decompose complex structures into
manageable components, facilitating their analysis
\citep{Neal_2020, McElreath_2020}. Second, they depict causal
relationships in a non-parametric way. This flexibility enables feasible
identification strategies without requiring specification of the types
of variables, the functional forms relating them, or the parameters of
those functional forms \citep[pp.~35]{Pearl_et_al_2016}.

@@

SCM~\ref{fig-cj03_scm} illustrates the inclusion of the first
theoretical principles into the conceptual-population model designed to
examine the impact of different teaching methods on students' writing
ability. This SCM outlines the relationship between the
conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
variables. The subscripts \(h\) and \(i\) label the students who
authored the texts (individuals), while the indices \(a\) and \(b\)
represent the compared texts (stimuli). The index \(j\) refers to the
judge, and \(k\) represents the judgment index, accounting for
experimental conditions where judges perform the same judgment multiple
times, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system
facilitates comparisons between texts written by the same student
\((h = i, a \neq b)\) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself \((h = i, a = b)\), as
such comparison lacks practical relevance within the CJ framework.

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:

\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
the number of stimuli available per individual. Likewise, each row of
\(JK\) associates a judge \(j\) with a judgment index \(k\).
Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
columns, where \(n_{K}\) indicates the number of judgments each judge
makes.

@@

It is easier to visualize the structure of these vectors and matrices by
considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
matrices described in equations (\ref{eq-mat11}) and (\ref{eq-mat12})
take the following form:

\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
}\end{equation}

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the latent variable representing judges' bias
\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes) causally influence the biases
associated with each text \(B_{JK}\), the second primary latent variable
of interest. Additionally, the branch accounts for idiosyncratic errors:
\(e_{J}\) captures variations in judges' bias unexplained by \(Z_{J}\),
while \(e_{JK}\) captures variations in judgments unexplained by
\(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
is an observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and
\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
latent vectors of length \(n_{J} \cdot n_{K}\).

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to a
\emph{conceptual population} of comparative data, encompassing all
repeated judgments from every judge for each pair of stimuli produced by
every individual in the population. Since researchers rarely meet this
assumption in practice, they must consider a more realistic scenario.
The extensions to the conceptual-population model, derived from this
realistic scenario, are described in the sample-comparison model.

@@

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. Expressly, we assume researchers work with
a data sample consisting of a \emph{limited number of judgments
\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\)}, all drawn from the conceptual population. In a
subsequent step, we assume that judges do not perform \emph{all possible
repeated comparisons} within the data sample. Instead, they conduct a
sufficient number of comparisons, \(n_{C}\), to ensure an accurate
estimation of the proportion \(P(B>A)\).

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the \emph{data sampling
process} using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:

\begin{equation}\phantomsection\label{eq-mat21}{
S_{I} = \begin{bmatrix}
i_{(1)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}
}\end{equation}

Where each element of \(S_{I}\) and \(S_{J}\) is a binary value
indicating the presence or absence of data elements in the \(I\) and
\(J\) vectors (see equation (\ref{eq-mat11})). Thus, the vectors
\(S_{I}\) and \(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements,
respectively. Specifically, we define the vector \(S_{I}\) as in
equation (\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).

\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is observed} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

Similarly, each element of \(S_{IA}\) and \(S_{JK}\) is a binary value
indicating the presence or absence of data rows in the matrices \(IA\)
and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{IA}\)
and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. Specifically, we define the vector \(S_{IA}\) as
in equation (\ref{eq-mat23}) and apply a similar definition to
\(S_{JK}\).

\begin{equation}\phantomsection\label{eq-mat23}{
ia_{(ia)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge. Additionally, they do not include the
second text from each individual and the third judgment from each judge.
Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
these vectors take the following structure:

\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all sampled students \((n^{s}_{I} \cdot n^{s}_{A})\), while
a similar sum in vector \(S_{JK}\) represents the total sampled
judgments across all sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Thus, the vector contains \(n\) elements, matching the
number of rows in the \(R\) matrix, and the sum of its elements
represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
}\end{equation}

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet[pp.~499-516]{McElreath_2020} and
\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} represents
the conceptual-population outcome \(O^{cp}_{R}\) as unobserved,
emphasizing that researchers cannot directly access this outcome due to
the sampling mechanism.

The DAG also depicts the \emph{sample design} vector \(S\) as a causal
factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
matrix (see equation (\ref{eq-mat12})), that is, this vector selects
\emph{all repeated comparisons a subset of judges makes for the subset
of stimuli produced by the sampled individuals}. This process ultimately
generates the sample-comparison outcome \(O^{sc}_{R}\) from the
conceptual-population outcome \(O^{cp}_{R}\).

@@

As in the previous section, we first define the \emph{comparison design}
using the binary vector variable \(C\) to facilitate the interpretation
of the sample-comparison model and integrate the comparison mechanism.

\begin{equation}\phantomsection\label{eq-mat26}{
C = \begin{bmatrix}
c_{(1)} \\
\vdots \\
c_{(c)} \\
\vdots \\
c_{(n)} \\
\end{bmatrix}
}\end{equation}

Where each element of \(C\) is a binary value indicating the presence or
absence of data rows in the \(R\) matrix (see equation
(\ref{eq-mat12})):

\[
c_{c} = \begin{cases} 
1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
\end{cases}
\]

Thus, the \emph{comparison design} vector \(C\) contains \(n\) elements
corresponding to the number of rows in the \(R\) matrix, and the sum of
its elements represents the selected comparisons, \(n_{C}\). This
definition assumes that judges do not perform \emph{all possible
repeated comparisons} within the data sample. Instead, they conduct a
sufficient number of comparisons, \(n_{C}\), to ensure the accurate
estimation of the proportion \(P(B>A)\) for each pair of stimuli
\citep[pp.~267]{Thurstone_1927b}.

@@

In summary, extending Thurstone's general form with causal inference
allows researchers to address BTL modeling issues, such as judge biases
(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
in trait estimation and hypothesis testing (see
Section~\ref{sec-theory-issue2}). However, the model does not address
concerns about the assumption of equal dispersion discussed in
Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
assumptions regarding the discriminal process distribution, it requires
a statistical model, which we will develop in the next section.

>>>>

Fortunately, SCMs and DAGs support identification analysis through two
key advantages. First, regardless of complexity, they can represent
various causal structures using only five fundamental building blocks
\citep[see][]{Neal_2020, McElreath_2020}. This feature allows
researchers to decompose complex structures into manageable components,
facilitating their analysis. Second, they depict causal relationships in
a non-parametric way. This flexibility enables feasible identification
strategies without requiring specification of the types of variables,
the functional forms relating them, or the parameters of those
functional forms \citep[pp.~35]{Pearl_et_al_2016}.

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
\(h\) and \(i\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) captures \emph{repeated
measures designs} \citep[pp.~366-376]{Lawson_2015}, accounting for
experimental conditions where a judge compares the same pair of stimuli
multiple times. Notably, the indexing system supports comparisons
between different texts written by the same student \((h = i;\)
\(a \neq b)\) and between texts written by distinct students
\((h \neq i;\) \(a = b \; \text{is permitted})\), each judged once or
repeatedly by all judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K)\).
However, it excludes cases where a judge compares a student's text to
itself, whether once or multiple times \((h = i;\) \(a = b;\)
\(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K})\), as such comparison lacks
practical relevance within the CJ framework. Here, \(n_{J}\) indicates
the total number of judges, and \(n_{K}\) denotes the number of repeated
judgments each judge performs.

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:
\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
n_{J} & 2 \\
\vdots & \vdots \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
the number of stimuli available per individual. Likewise, each row of
\(JK\) associates a judge \(j\) with a judgment index \(k\).
Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
columns, where \(n_{K}\) indicates the number of repeated judgments each
judge makes.

@@

It is easier to visualize the structure of these vectors and matrices by
considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
matrices described in equations (\ref{eq-mat11}) and (\ref{eq-mat12})
take the following form:
\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
}\end{equation}

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). This
branch also shows how \(B_{J}\), along with \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes), and judgments' idiosyncratic errors
\(e_{JK}\) causally influence the biases associated with each text
\(B_{JK}\), the second primary latent variable of interest. The error
\(e_{JK}\) captures variations in judgments unexplained by \(B_{J}\) or
\(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with \(n_{J}\) rows
and \(q_{J}\) columns (variables), and both \(e_{J}\) and \(B_{J}\) are
latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\) is an observed
matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\) columns
(variables), while \(e_{JK}\) and \(B_{JK}\) are latent vectors of
length \(n_{J} \cdot n_{K}\).

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issue1b} and Section~\ref{sec-theory-issue2}.
These enhancements include accounting for judges' biases and integrating
hierarchical structural components. Nevertheless, despite its promise of
enhancing measurement accuracy and precision, the model still depends on
the unrealistic assumption that researchers have access to a
\emph{conceptual population} of comparative data. Since researchers
rarely meet this assumption in practice, they must consider a more
realistic scenario. The extensions to the conceptual-population model,
derived from this realistic scenario, are described in the
sample-comparison model.

@@

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. Expressly, it assumes researchers work with
a data sample consisting of a limited number of repeated judgments
\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\), all drawn from the conceptual population. In a
subsequent step, it assumes that judges do not perform \emph{all
possible stimuli comparisons} within the data sample. Instead, they
conduct a sufficient number of stimuli comparisons, \(n_{C}\), to ensure
an accurate estimation of the proportion \(P(B>A)\).

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the \emph{data sampling
process} using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:
\begin{equation}\phantomsection\label{eq-mat21}{
S_{I} = \begin{bmatrix}
i_{(1)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}
}\end{equation}

Where each element of \(S_{I}\) and \(S_{J}\) is a binary value
indicating the presence or absence of data elements in the \(I\) and
\(J\) vectors (see equation (\ref{eq-mat11})). Thus, the vectors
\(S_{I}\) and \(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements,
respectively. Specifically, we define the vector \(S_{I}\) as in
equation (\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).
\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is observed} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

Similarly, each element of \(S_{IA}\) and \(S_{JK}\) is a binary value
indicating the presence or absence of data rows in the matrices \(IA\)
and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{IA}\)
and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and \(n_{J} \cdot n_{K}\)
elements, respectively. Specifically, we define the vector \(S_{IA}\) as
in equation (\ref{eq-mat23}) and apply a similar definition to
\(S_{JK}\). \begin{equation}\phantomsection\label{eq-mat23}{
ia_{(ia)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose researchers exclude the second
student and the third judge. Additionally, they do not include the
second text from each individual and the third judgment from each judge.
Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
these vectors take the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\). In other words, researchers can only observe texts and
judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all sampled students \((n^{s}_{I} \cdot n^{s}_{A})\), while
a similar sum in vector \(S_{JK}\) represents the total sampled repeated
judgments across all sampled judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample mechanism} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Thus, the vector contains \(n\) elements, matching the
number of rows in the \(R\) matrix, and the sum of its elements
represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.
\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
}\end{equation}

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet[pp.~499-516]{McElreath_2020} and
\citet[pp.~6]{Deffner_et_al_2022}, DAG~\ref{fig-cj12_dag} represents the
conceptual-population outcome \(O^{cp}_{R}\) as unobserved, emphasizing
that researchers cannot directly access this outcome due to the sampling
mechanism. The DAG also depicts the \emph{sample design} vector \(S\) as
a causal factor influencing the sample-comparison outcome
\(O^{sc}_{R}\). A square encloses this vector, indicating that it is a
conditioned variable. In this context, \emph{conditioning} means
restricting the statistical analysis to a specific subset of the
population based on the values of a conditioned variable
\citep[pp.~32]{Neal_2020}, in this case \(S\). Essentially, \(S\) serves
as the sample equivalent for the \(R\) matrix (see equation
(\ref{eq-mat12})), that is, this vector selects \emph{all repeated
judgments a subset of judges makes for the subset of stimuli produced by
the sampled individuals}. This process ultimately generates the
sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
outcome \(O^{cp}_{R}\).

@@

As in the previous section, to facilitate the interpretation of the
sample-comparison model, we first define the \emph{comparison mechanism}
using the binary vector variable \(C\) as follows:
\begin{equation}\phantomsection\label{eq-mat26}{
C = \begin{bmatrix}
c_{(1)} \\
\vdots \\
c_{(c)} \\
\vdots \\
c_{(n)} \\
\end{bmatrix}
}\end{equation}

Where \(C\) contains \(n\) elements corresponding to the number of rows
in the \(R\) matrix, with each element being a binary value indicating
the presence or absence of data rows in the \(R\) matrix (see equation
(\ref{eq-mat12})): \[
c_{(c)} = \begin{cases} 
1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
\end{cases}
\]

This definition assumes that the subset of judges \emph{do not} perform
all repeated judgments for the subset of stimuli produced by the sampled
individuals. Instead, they conduct a ``sufficient number of
comparisons'', \(n_{C}\), to ensure the accurate estimation of the
proportion \(P(B>A)\) for each pair of stimuli
\citep[pp.~267]{Thurstone_1927b}.

@@

In summary, extending Thurstone's general form using Causal inference
allows researchers to address BTL modeling issues, such as judge biases
(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
in trait estimation and hypothesis testing (see
Section~\ref{sec-theory-issue2}). However, the model does not address
concerns about the assumption of equal dispersion discussed in
Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
assumptions regarding the discriminal process distribution, it requires
a statistical model, which we will develop in the next section.
